Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-09-14 iso-int/rc-iso-int.tex
_

\paragraph*{Steps of the Construction.}

Although trivial, the previous example captures the fundamental aspects of
our construction. In general, our starting point is an abstract machine in
which every rewrite step is logically reversible. The formal criterion of
logical reversibility in this setting is captured by Abramsky's
\emph{bi-orthogonal automata}~\cite[Sec.~2]{abramsky2005structural} which is
summarized as:
\begin{enumerate}
\item
  Machines must have distinguishable start and stop states. There may
  be more than one valid start state and more than one valid stop
  state, however.
\item Each valid machine state must match a unique pattern on the
  left-hand side and right-hand side of the `{{|-->}}' relation.
\item Every reduction step must be computable (in {{langRevT}}), must
  not drop or duplicate pattern variables and must be logically
  reversible --- i.e. it must be possible to reduce from right to
  left.
\end{enumerate}
Given such a machine the process of encoding the machine in {{langRevT}}
consists of the following steps:
\begin{enumerate}
\item Expand the input to expose enough structure to distinguish the
  left-hand side of each machine state;
\item Expand the output to expose enough structure to distinguish the
  right-hand side of each machine state;
\item Shuffle matching input terms to output terms, inserting any appropriate
  mediating computations. 
\item Slide the two sections to expose the start and stop state and
  introduce a {{trace}} to iterate the construction. 
\end{enumerate}

@@

>>>>

\subsection{Steps of the Construction.}

Although trivial, the previous example captures the fundamental aspects of
our construction. In general, our starting point is an abstract machine in
which every rewrite step is logically reversible. The formal criterion of
logical reversibility in this setting is captured by Abramsky's
\emph{bi-orthogonal automata}~\cite[Sec.~2]{abramsky2005structural} which is
summarized as:
\begin{enumerate}
\item 
 Machine states are described as tuples whose components are algebraic
 data types. In the above example we used only the {{nat}} data type
 and tuples of the form {{<n, m>}}.
\item
  Machines must have distinguishable start and stop states. There may
  be more than one valid start state and more than one valid stop
  state, however.

\item Each valid machine state must match a unique pattern on the
  left-hand side and right-hand side of the `{{|-->}}' relation.

\item Every reduction step must be computable (in {{langRevT}}), must
  not drop or duplicate pattern variables and must be logically
  reversible --- i.e. it must be possible to reduce from right to
  left.
\end{enumerate}
Given such a machine the process of encoding the machine in {{langRevT}}
consists of the following steps:
\begin{enumerate}
\item Expand the input to expose enough structure to distinguish the
  left-hand side of each machine state;
\item Expand the output to expose enough structure to distinguish the
  right-hand side of each machine state;
\item Shuffle matching input terms to output terms, inserting any appropriate
  mediating computations. 
\item Slide the two sections to expose the start and stop state and
  introduce a {{trace}} to iterate the construction. 
\end{enumerate}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parity Translation }
\label{sec:parity-int}

Deriving {{langRevT}} combinators from abstract machines can sometimes
be used as an efficient indirect way of programming in {{langRevT}}.
Say we wished to build a {{langRevT}} combinator to map {{nat}} to a
parity encoded version {{par = @@x.1+(1+x)}}. Since {{par}} is not a
fixed unfolding of {{nat}}, it is not immediately apparent how such a
combinator can be constructed.

Instead of directly programming in {{langRevT}}, one can derive the
required {{nat<->par}} combinator by first constructing an abstract
machine that maps {{nat}} to {{par}}. We first express {{nat}} and
{{par}} algebraically and then develop a logically reversible abstract
machine expressing the translation:

% Algebraically, we may write:

%% %subcode{bnf} include main
%% % Numbers, n, m = 0 | n + 1
%% % Parity, par = zero | one | plus2 par

%% While it is apparent that there is a natural one to one correspondence
%% between their values, it is non-obvious how this can be expressed as a
%% {{langRevT}} isomorphism. This map can be expressed in terms of the
%% algebraic type definitions as shown below:

%% %subcode{opsem} include main
%% %! columnStyle = ||c|r||
%% % 0 & even
%% % 1 & odd
%% % 2 & A even
%% % 3 & A odd
%% % 4 & A (A even)
%% % ... &  ...

%% \indent
%% and in terms of their {{langRevT}} type encodings as shown below:

%% %subcode{opsem} include main
%% %! columnStyle = ||c|c||
%% % <left ()> & <left ()>
%% % <right <left ()>> & <right left ()>
%% % <right <right <left ()> > > & <right right <left ()>>
%% % <right <right <right <left ()>> >> & <right right <right left ()>>
%% % <right <right <right <right <left ()>> >> > & <right right <right right <left ()> > >
%% % ... &  ...

% \noindent

%subcode{bnf} include main
% Numbers, n, m = 0 | n + 1
% Parity, par = even | odd | A par
%

\noindent
Machine States:

%subcode{bnf} include main
% Machine states = <nat, nat> | {[parity, nat]}
% Start state = <n , 0>
% Stop state =  {[par, 0]}

\noindent
Machine Reductions:
%subcode{opsem} include main
%! columnStyle = rclr
% <0, m> &|-->& {[even, m]}
% <1, m> &|-->& {[odd, m]}
% <(n+1)+1, m> &|-->& <n, m+1>
% {[par, m+1]} &|-->& {[A par, m]}

The construction of the {{langRevT}} combinator is
straightforward and proceeds as before. The partial combinator
representing the wiring of machine reductions is shown below.

\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/nat-parity.pdf}
}
\end{center}

\noindent
We thank Fritz Henglein for the motivation underlying this example.
